---
title: "Getting Started with dcsbm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dcsbm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5)
```

## Motivation

`dcsbm` fits a Bayesian weighted degree-corrected stochastic block model
(DC-SBM) to a symmetric weighted network: nodes are partitioned into `K`
blocks, and the expected edge weight between two nodes depends on which pair
of blocks they belong to, plus a per-node degree-correction term. Unlike a
standard SBM, edge weights are continuous (Gaussian), not just present/absent.

The package also supports an optional **edge-level covariate effect** that is
*localized* to each block pair rather than shared across the whole network —
useful when you suspect a covariate's association with edge weight differs
by which communities the two endpoints belong to.

This vignette walks through simulating a network with known structure,
fitting the base model, adding a covariate effect, choosing `K`, and using
the `reallocate` move to improve chain mixing.

```{r setup}
library(dcsbm)
```

## Simulating a network

We simulate a 24-node network with three well-separated blocks: strong
positive within-block weights, negative between-block weights.

```{r simulate}
set.seed(42)
n <- 24
K_true <- 3
z_true <- rep(1:K_true, each = n / K_true)

G_true <- matrix(-1.5, K_true, K_true)
diag(G_true) <- 2.5

W <- matrix(0, n, n)
for (i in seq_len(n)) {
  for (j in seq_len(n)) {
    if (i != j) W[i, j] <- rnorm(1, G_true[z_true[i], z_true[j]], 0.25)
  }
}
W <- (W + t(W)) / 2
```

## Fitting the base model

With `K` known (here, the true value), `fit_dcsbm()` runs a Gibbs sampler and
returns posterior draws and summaries.

```{r fit-base}
fit <- fit_dcsbm(W, K = K_true, n_iter = 400, burn = 200, thin = 2, seed = 1, verbose = FALSE)
print(fit)
```

The MAP block assignment recovers the planted structure up to label
permutation:

```{r check-recovery}
table(true = z_true, estimated = fit$summary$z_hat)
```

Diagnostic plots are available via `plot()`:

```{r plot-base}
plot(fit, type = "loglik")
plot(fit, type = "heatmap")
```

## Adding a localized edge covariate

Suppose we also observe a continuous edge-level covariate `X` (e.g., a
pairwise similarity score) whose association with edge weight we want to
estimate separately for every pair of blocks, rather than assuming one
global slope.

```{r fit-covariate}
X <- matrix(0, n, n)
X[upper.tri(X)] <- runif(sum(upper.tri(X)), 0, 1)
X <- X + t(X)

fit_x <- fit_dcsbm(W, K = K_true, X = X, n_iter = 400, burn = 200, thin = 2, seed = 1, verbose = FALSE)
round(fit_x$summary$beta_post_mean, 2)
```

`beta_post_mean` is a `K x K` matrix of posterior mean covariate slopes, one
per block pair — this is the localized effect the package is built around.

## Choosing K

When the true number of blocks is unknown, `select_K()` sweeps a grid of `K`
values and reports fit criteria for each, so you can look for an elbow in
mean log-likelihood or check for degenerate (empty/tiny) blocks. If a
particular `K` is too large for the data to support — every post-burn
iteration hits an empty block — that row is filled with `NA` and a warning
is raised, but the rest of the grid still completes. With only 24 nodes and
3 well-separated true blocks, `K = 4` is a case in point below.

```{r select-k}
grid <- select_K(W, K_grid = 2:4, n_iter = 400, burn = 200, thin = 2, verbose = FALSE)
grid
plot(grid)
```

## The reallocate move

Single-node Gibbs updates only move one node at a time, so a chain can get
stuck when several nodes would only want to swap blocks together as a group.
The optional `reallocate = TRUE` move jointly redistributes node memberships
between two existing blocks in a single Metropolis-Hastings step, which can
substantially improve mixing (in one real-data comparison, effective sample
size of the log-likelihood trace improved roughly 24x on an otherwise
identical chain).

```{r reallocate}
fit_ra <- fit_dcsbm(W, K = K_true, reallocate = TRUE,
                     n_iter = 400, burn = 200, thin = 2, seed = 1, verbose = FALSE)
fit_ra$summary$realloc_accept_rate
```

## Next steps

See `?fit_dcsbm` for the full list of prior-variance and sampler-control
arguments, and `?select_K` for details on how failed fits (e.g. a `K` that
never saves a draw with all blocks occupied) are handled in the grid sweep.
